---
title: "Forest Fire"
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(dplyr)
library(gganimate)
source("./helpers/forest_fire.R")
source("./helpers/grid_plot.R")
```


You can even model forest fires with agent-based modeling.
Here is how.

Take a square grid.
Each cell has one of four distinct states:

* `white` = The cell is empty
* `green` = There is a tree on the cell
* `red` = There is a burning tree on the cell
* `dark red` = There is a burnt tree on the cell

To set up the model, we distribute trees (= green cells) over the grid and set a certain number of initial trees on fire.
The effects of the model are most demonstrative, when we set the first column on fire.
This means that all trees on the left of the board are on fire.

```{r}
forest <- create_forest(50, 0.7)

forest_fire_simulation <- forest_fire_game(forest, 200)

color_scheme <- c(`0` = "white", `1` = "green", `2` = "red", `3` = "firebrick")

forest_fire_simulation %>% 
  grid_plot(2, color_scheme)
```


In one model step, each burning tree sets all non-burning trees in its **Von Neumann neighborhood** on fire.
All burning trees burn out (i.e., the become "burnt" or like described above `dark red`).

One parameter that we can manipulate is the density of trees on the board.
Let's have a look at an example with a tree density of 0.7:

```{r}
forest_fire_simulation <- forest_fire_game(forest, 200)

color_scheme <- c(`0` = "white", `1` = "green", `2` = "red", `3` = "firebrick")

anim <- forest_fire_simulation %>% 
  grid_plot(2, color_scheme) +
  transition_states(step, transition_length = 1, state_length = 10)

animate(anim, nframes = 201, renderer = gifski_renderer())
```






























