---
title: "Conway's Life"
output: distill::distill_article
bibliography: bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
library(dplyr)
```

This game has first been described in a public newsletter by Martin Gardner [@gardner1970mathematical].

Take a square grid of sufficient size
Let' start with something small, say 10x10.
Randomly assign each cell one of two states: alive or dead.

This is what we have so far:

```{r}
axis_size <- 10

df <- data.frame(
  x = rep(seq(1, axis_size), each = axis_size),
  y = rep(seq(1, axis_size), times = axis_size),
  state = sample(c("alive", "dead"), replace = TRUE, axis_size^2)
)

color_scheme <- c("alive" = "black", "dead" = "white")

df %>% 
  ggplot(aes(x = x, y = y, fill = state)) +
  geom_point(size = 13, shape = 22, color = "lightgrey") +
  scale_fill_manual(values = color_scheme) +
  coord_fixed() +
  theme(
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none",
    panel.background = element_rect(fill = "white"),
    panel.grid = element_blank()
  ) +
  NULL
```


We will now update the automaton one step at a time according to the following rules [@gardner1970mathematical]:

* Survivals: Each alive cell with 2 or 3 neighbors survives into the next generation.
* Deaths: Each alive cell that has less than 2 neighbors dies from isolation. Each alive cell with 4 or more neighbors dies from overpopulation.
* Births: Each dead cell with exactly 3 neighbors becomes alive.

If we talk about neighbors, we are referring to the Moore neighborhood.
Note that we apply these rules to all cells **simultaneously**.

So far so good.
Seems innoccuous, right?

#### Task

What would you expect to happen when we play this setup out over many generations?


In fact, this cellular automaton, called *Life*, displays some very complex dynamics, and, frankly, some really fancy ones.

Let's have a look at an example.

[ANIMATION OF LIFE]


TODO:

* gliders, spaceships
* a bit of history
* intuitive notion of emergence







#### Project

Implement *Life* in Julia.